\thetitle

### 2020-07-08 writing/rational.tex
rational programmer touch ups

Based on this idea, a metric for migratory paths is a concise way to
evaluate the effectiveness of a
rational programmer. The components of each program create a
\emph{debugging scenario
space}, like the one introduced by~\citet{tfgnvf-popl-2016}. This space consists
of variants of the program, dubbed \emph{scenarios}, that differ in their
type specifications. In the setting of migratory typing,\footnote{The method can
be adapted to so-called true gradual typing systems~\cite{svcb-snapl-2015} in a
straightforward manner.} each scenario differs from another in which
components are typed. Formally, a program $\system$ is a set of components
$\set{\component}$, and a scenario $\conf$ is a subset of
$\set{\component}$ . These scenarios of $\system$ are ordered by the subset
relation and thus form a lattice $\lattice{\system}$ with $2^{\size{\system}}$
elements.  The bottom scenario of $\lattice{\system}$ is always $\emptyset$
and the top one is $\system$ itself. In between are all the mix-typed
scenarios.

With these definitions, the actions of the rational programmer map to an
ascending chain---dubbed a \emph{trail}---in $\lattice{\system}$. 
Intuitively, with each action the rational programmer  reduces the current
scenario to an ``easier'' one to debug. The easiest scenarios are those
that the type checker rejects outright. Assuming the types of the
program are correct, if the typechecker signals a type error, the error
pinpoints an impedance mismatch between the broken code of a component 
and its type. Hence the programmer knows which component to fix. 
The starting
scenario $\conf_0$ of a trail---also referred to as the \emph{root} (of the
trail)---is the scenario that the programmer attempts to debug initially. Any other
configuration $\conf_{i+1}$ in the trail differs from its predecessor
$\conf_{i}$ in that more 
components compared to  $\conf_{i}$ become typed. The rational programmer 
decides which are these components based on the feedback it receives from
the gradual typing system when running $\conf_{i}$.

\footnote{Blame can result either directly from a run time type
check or, in the absence of blame, from an exception that we track back to the
component that causes it with the help of the stack trace.}

The next three subsections specialize the idea to the various semantics of
gradual typing. Once this machinery is in place, we use the fourth subsection to
state the precise experimental research questions that correspond to the
high-level question on the first page and to explain our experimental
process. The final subsection extends the development with a discussion of
how we can measure the rational programmer's effort.

@@

In natural Typed Racket, rational programmers rely on blame when they
attempt to debug a mix-typed programmer; they add type specifications to
the component blame points. In terms of scenarios and trails, this
translates to extending a trail with a new scenario that adds to the
latest scenario from the trail the component that latest scenario
blames.  We call trails constructed in this manner \emph{natural blame
trails} and they capture the essence of the natural rational programmer
:\footnote{We use trail and rational programmer interchangeably from now
on.}

\begin{quote}
\it 
A natural blame trail
is a sequence of scenarios $\conf_0,...\conf_n$ of a program
$\system$  such that for all $0 \leq i \leq n - 1$, $\conf_i \subset \conf_{i+1}$ and
  $\conf_{i+1} \setminus \conf_i =\{\blame{\system, \conf_i}\}$ where
  $\blame{\system, \conf}$ denotes the component that $\conf$ blames.
\end{quote}

Natural rational programmers migrate faulty programs along natural blame
trails in order to locate bugs. As we mention above, assuming that the types 
they add during migration are correct, when the buggy component of a
program migrates from untyped to typed, the typechecker detects the issue
at compile time. Hence, trails that end at an ill-typed scenarios successfully 
pinpoint the location of bugs:
In formal terms:
\begin{quote}
\it 
A natural blame trail $\conf_0,...\conf_n$ in a lattice $\lattice{\system}$ is
\emph{successful}  iff its last scenario $\conf_n$ does not type check.
  \end{quote}
  In juxtaposition:
\begin{quote}
\it 
A natural blame trail $\conf_0,...\conf_n$ in a lattice $\lattice{\system}$ is
\emph{failing}  iff $\conf_n$ type checks and the trail cannot be extended
  further. 
 \end{quote}
\noindent 
That is failing natural blame trails are those that have not reached a scenario that reveals
the bug at compile time and their last scenario does not produce
blame. Thus the rational programmer receives no hints from the gradual type
system about how to further migrate the faulty program in search for the
bug.

 We can now give a precise definition of what it
means for blame to help the rational programmer:
\begin{quote}
\it
  Natural blame is always useful for a program $\system$ iff 
  $\lattice{\system}$ contains no failing natural blame trails.

   Natural blame is sometimes useful for a program $\system$ iff 
  $\lattice{\system}$ contains successful natural blame trails.

   Natural blame is never useful for a program $\system$ iff 
  $\lattice{\system}$ contains no successful natural blame trails.

These definitions allow a rational programmer to gauge whether it pays off to heed blame
when debugging \system. 
However, they do not answer whether blame is a critical piece of 
the rational programmer's process.  For instance, 
typing a random component   could be equally useful as typing the blamed one. 
To account for this situation, we devise a new
mode of the natural rational programmer that 
 follows a random migration process:
\begin{quote}
\it 
A natural random trail
is a sequence of scenarios $\conf_0,...\conf_n$ of a program
$\system$  such that for all $0 \leq i \leq n - 1$, $\conf_i \subset \conf_{i+1}$ and
$\conf_{i+1} \setminus \conf_i = \{\random{\system, \conf_i}\}$
where $\random{\system, \conf}$ denotes a randomly selected component of $\system$
that is not in $\conf_i$.
\end{quote}
\indent
Similar to natural blame we determine how successful the random
migration is based on the success of natural random trails: 
\begin{quote}
\it
  Natural random is always useful for a program $\system$ iff 
  $\lattice{\system}$ contains no failing natural random trails.

   Natural random is sometimes useful for a program $\system$ iff 
  $\lattice{\system}$ contains successful natural random trails.

   Natural random is never useful for a program $\system$ iff 
  $\lattice{\system}$ contains no successful natural random trails.

Putting all these pieces together, we can ask whether 
blame adds value over random migration: 
\begin{quote}
\it
  Natural blame is more successful than natural random for $\system$ if 
  either (i) natural blame is always successful for $\system$ while natural random 
  is sometimes or never successful for $\system$, or (ii) 
  natural blame is sometimes successful for $\system$ while natural random 
  is never successful for $\system$.

  Natural blame adds value to natural for a program $\system$ over pure chance iff 
  natural blame is more successful than natural random.
\end{quote}

Even if the success of the natural rational programmer is not by pure
chance, it could well be because the natural semantics perform runtime type checks at
the right time. In other words, one rational programmer that 
heeds blame could be equally successful as one that only uses information from the exceptions of failed 
runtime type checks. Similar to before, in response we
define the exception mode of the rational programmer:
\begin{quote}
\it 
A natural exception trail
is a sequence of scenarios $\conf_0,...\conf_n$ of a program
$\system$ such that for all $0 \leq i \leq n - 1$, $\conf_i \subset \conf_{i+1}$ and
$\conf_{i+1} \setminus \conf_i = \{\error{\system, \conf_i}\}$
  where $\error{\system, \conf}$ denotes the component that triggers a
an exception of the underlying language.
\end{quote}
\noindent
We use the new mode to ask whether blame adds value over both exceptions and pure
chance in natural:
\begin{quote}
\it
  Natural blame adds value to natural for a program $\system$ over pure chance and
  natural exceptions iff natural blame is more successful than both natural
  random and natural exception.
\end{quote}
\noindent
If blame does add value to natural for a program $\system$ 
over pure chance and exceptions then we can finally
deduce that, for debugging $\system$, blame is a valuable feature of the
natural semantics.

@@

 Since gradually typed languages with erasure semantics do not come with
 blame assignment, a rational programmer can only hope that the underlying
 safety checks and their exceptions are helpful.  Thus, adapting the
 definitions related to exception and random trails from natural to
 erasure, we ask whether the effect of erasure exceptions in debugging a
 program $\system$ goes beyond that of pure chance:

@@

In contrast to erasure, transient does come with blame. As we discuss in
section~\ref{sec:landscape} though, distinctly from natural blame,
transient may blame more than one components. This peculiarity of 
transient raises the question of how the rational programmer should react
when the language produces a blame set. Our answer is that the rational
programmer has at least two reasonable options.  The first option is the rational programmer types
all blamed components at once.  In the second option, the rational
programmer selects the component that is added to the blame set first and types
only that --- after all the types of this first component should be able
to  detect an impedance mismatch earlier in the evaluation of a program. 
As before, the two modes give rise to two different notions of trail:
\begin{quote}
\it 
A  transient all blame trail
is a sequence of scenarios $\conf_0,...\conf_n$ of a program
$\system$ where for all $0 \leq i \leq n - 1$, $\conf_i \subset \conf_{i+1}$ and
~$\conf_{i+1} \setminus \conf_i = \mblame{\system, \conf_i}$
where $\mblame{\system, \conf}$ denotes the set of components that $\conf$
blames under transient semantics; 
\end{quote}
\noindent
and
\begin{quote}
\it 
A transient firstborn blame trail
is a sequence of scenarios $\conf_0,...\conf_n$ of a program
$\system$ where for all $0 \leq i \leq n - 1$, $\conf_i \subset \conf_{i+1}$ and
  $\conf_{i+1} \setminus \conf_i = \first{\mblame{\system, \conf_i}}$
  where $\first{\mblame{\system, \conf}}$ is the first component 
  the transient semantics add to the blame set for $\conf$.
\end{quote}
\noindent As for natural we use the transient all blame trail transient
firstborn, transient erasure and transient random trails to define when
transient all blame and transient firstborn add value to transient over
exceptions and pure chance: 
\begin{quote}
\it 
Transient all blame adds value to transient for a program $\system$ over
  pure chance and transient exceptions iff transient all blame 
  is more successful than both transient random and transient exception.

@@

\begin{figure}
\center
{\begin{tabular}[ht]{l|c|c|c}
%% ---------------------------------------------------------------------------------------------------------------
                        & {\bf Natural}        & {\bf Transient}          & {\bf Erasure} \\ \hline 
%% ---------------------------------------------------------------------------------------------------------------
{\bf Blame}             &       $Q_1$          &                          &               \\
{\bf All blame}         &                      &     $Q_2$                &               \\
{\bf Firstborn blame}   &                      &     $Q_3$                &               \\
{\bf Exception}         &       $Q_1$/$Q_*$    &     $Q_2$/$Q_3$/$Q_*$    &      $Q_*$    \\
{\bf Random}            &       $Q_1$/$Q_*$    &     $Q_2$/$Q_3$/$Q_*$    &      $Q_*$    \\
\end{tabular}}
  \caption{ Experimental questions and rational programmer modes.}
  \label{fig:experiment-outline}
\end{figure}

@@

In detail, to answer $Q_1$ for a mutant, we have to 
calculate and log for each natural blame, natural exception and natural
random trail 
in the scenario lattice of the mutant whether it is successful or failing. 
Hence, we first construct the mutant's scenario
lattice and we determine the scenario of the lattice that can detect the fault in the
program. This can happen either because Typed Racket's type checker
outright rejects the scenario or because natural Typed Racket's runtime 
performs a type check that fails. In either
case,  each of these scenarios is the root of a trail that we attempt
extend according to the natural blame rational programmer. 
If no scenarios can be added to the trail\footnote{All trails
reach that point since the lattices are finite.}, we check if the
last scenario of the trail type checks or not. If it does this
natural blame trail
is successful, otherwise it is failing. We repeat the same process for the
roots two more times; once following the natural exception rational programmer and 
once the natural random rational programmer. Finally, based on the
collected data, we
calculate the success/failure results for the trails to determine whether
 natural blame is more successful than  natural exceptions and pure chance
 and we obtain the answer for $Q_1$.
The process is analogues for the other questions employing their corresponding
modes of the rational programmer.

>>>>

Based on this idea, a metric for migratory paths is a concise way to
evaluate the effectiveness of a rational programmer. The components of
each program create a \emph{debugging scenario space}, a lattice like
the one introduced by~\citet{tfgnvf-popl-2016} for performance
evaluations. This space consists of variants of the program, dubbed
\emph{scenarios}, that differ in their type specifications. In the
setting of migratory typing,\footnote{The method can be adapted to
so-called true gradual typing systems~\cite{svcb-snapl-2015} in a
straightforward manner.} each scenario differs from another in which
components are typed. Formally, a program $\system$ is a set of
components $\set{\component}$, and a scenario $\conf$ is a subset of
$\set{\component}$ . These scenarios of $\system$ are ordered by the
subset relation and thus form a lattice $\lattice{\system}$ with
$2^{\size{\system}}$ elements.  The bottom scenario of
$\lattice{\system}$ is always $\emptyset$ and the top one is $\system$
itself. In between are all the mix-typed scenarios.

With these definitions, the actions of the rational programmer map to
an ascending chain---dubbed a \emph{trail}---in $\lattice{\system}$.
The starting scenario $\conf_0$ of a trail---also referred to as the
\emph{root} (of the trail)---is the scenario that the programmer
attempts to debug initially. The easiest scenarios are those that the
type checker rejects outright, because a type error pinpoints an
impedance mismatch between the broken code of one component and its
typed interface to others. At this point, the programmer has
identified the broken component. If the scenario type-checks, the
programmers runs the program until it raises an exception.  The
rational programmer uses the blame assignment\footnote{Blame can
result either directly from a run time type check or, in the absence
of blame, from an exception that we track back to the component that
causes it with the help of the stack trace.} to decide which
component(s) to equip with types. This choice constructs scenario
$\conf_{i+1}$ from $\conf_{i}$ and implies the creation of an
ascending chain.  Intuitively, with each step along the chain, the
rational programmer reduces the current scenario to an ``easier'' one
to debug.

% The successor of a scenario on a trail has more typed component;
% formally, configuration $\conf_{i+1}$ on the trail differs from its
% predecessor $\conf_{i}$ in that more components are typed.

The next three subsections specialize the trail-scenario idea to the
various semantics of gradual typing. Once this machinery is in place,
we use the fourth subsection to state the precise experimental
research questions that correspond to the high-level question on the
first page and to explain our experimental process. The final
subsection extends the development with a discussion of how we can
measure the rational programmer's effort along the trail.

@@

The natural semantics assigns blame to exactly one boundary.  A blame assignment
has the following specific meaning: the typed component may make incorrect type
assumptions about the untyped component in its interface, or the correct
interface exposes a bug in the untyped component. Our setup rules out the first
alternative (but see section~\ref{sec:conclusion}), and therefore the rational
programmer extends the trail to a scenario that swaps out the untyped component
for its typed counterpart.

We can turn this description into a formal definition of the trail. 
\begin{quote}
\it A \emph{natural blame trail} is a sequence of scenarios $\conf_0,...\conf_n$ of
a program $\system$ such that for all $0 \leq i \leq n - 1$, $\conf_i \subset
\conf_{i+1}$ and $\conf_{i+1} \setminus \conf_i =\{\blame{\system, \conf_i}\}$ where
$\blame{\system, \conf}$ denotes the component that $\conf$ blames.
\end{quote}

When the buggy untyped component of a program is replaced by the typed counterpart,
the type checker fails if this component causes the impedance mismatch. Hence, a
trail that end at an ill-typed scenarios successfully pinpoints the location of
bugs. 
\begin{quote}
\it A natural blame trail $\conf_0,...\conf_n$ in a lattice $\lattice{\system}$ is
\emph{successful} iff its last scenario $\conf_n$ does not type check.  A natural
blame trail $\conf_0,..,\conf_n$ in a lattice $\lattice{\system}$ is \emph{failing}
iff $\conf_n$ type checks and the trail cannot be extended further.
\end{quote}
That is failing natural blame trails are those that have not reached a scenario that
reveals the bug at compile time and their last scenario does not produce blame. Thus
the rational programmer receives no hints from the gradual type system on how to
continue the search for the bug.

Equipped with these preliminary definitions, we can precisely define what it means
for blame to add value to a gradual type system in a specific setting. 
\begin{quote}
\it
Given a program $\system$, natural blame is\\
(1) \emph{always useful} iff $\lattice{\system}$ contains no failing natural blame trails.\\
(2) \emph{sometimes useful} iff $\lattice{\system}$ contains successful natural blame trails.\\
(3) \emph{never useful} iff $\lattice{\system}$ contains no successful natural blame
trails.\\
\end{quote}

While these definitions describe how a rational programmer might gauge whether it
pays off to heed blame assignments during debugging, they do not answer whether
blame is a critical piece of the rational programmer's process.  For instance,
typing a random component might be as useful as typing the blamed one.

To account for this situation, we devise a new mode of the natural rational
programmer that follows a random migration process:
\begin{quote}
\it A natural random trail is a sequence of scenarios $\conf_0,...\conf_n$ of a
program $\system$ such that for all $0 \leq i \leq n - 1$, $\conf_i \subset
\conf_{i+1}$ and $\conf_{i+1} \setminus \conf_i = \{\random{\system, \conf_i}\}$
where $\random{\system, \conf}$ denotes a randomly selected component of $\system$
that is not in $\conf_i$.
\end{quote}
Similar to natural blame we determine how successful the random
migration is based on the success of natural random trails: 
\begin{quote}
\it
Given a program $\system$, natural random blame is\\
(1) \emph{always useful} iff $\lattice{\system}$ contains no failing natural random blame trails.\\
(2) \emph{sometimes useful} iff $\lattice{\system}$ contains successful natural random blame trails.\\
(3) \emph{never useful} iff $\lattice{\system}$ contains no successful natural
random blame trails.\\
\end{quote}

Putting the pieces together, we can now define that blame assignments adds value in
the context of a natural semantics if blame-based migration is better than random
migration.
\begin{quote}
\it Natural blame adds value to natural for a program $\system$ over pure chance iff
  natural blame is more successful than natural random. And, natural blame is more
  successful than natural random for $\system$ if either natural blame is always
  successful for $\system$ when natural random is sometimes or never successful, or
  natural blame is sometimes successful for $\system$ when natural random is never
  successful.
\end{quote}

Even if the success of the natural rational programmer is not by pure chance, it
could well be because the natural semantics perform runtime type checks at the same

@@

 Since gradually typed languages with erasure semantics do not come with
 blame assignment, a rational programmer can only hope that the underlying
 safety checks and their exceptions are helpful.  Thus, adapting the
 definitions related to exception and random trails from natural to
 erasure, we characterize when the effect of erasure exceptions in debugging a
 program $\system$ goes beyond that of pure chance.

@@

The transient semantics assigns blame to a series of components. In this setting, a
blame assignment has the following meaning: the witness value to the impedance
mismatch crossed the boundaries between neighbors in the sequence, and each crossing
checked a portion of the value's type shape.

This ambiguity in transient blame raises the question of how the rational programmer
should react when the language produces a blame set. Our answer is that the rational
programmer has at least two reasonable options.  The first option is to type all
blamed components at once. The second one is to select the component that is added
to the blame set first and assign types to only that one---after all the types of
this first component should be able to detect an impedance mismatch earlier in the
evaluation of a program than the later ones.

These two modes of rationalizing give rise to two different notions of trail.
\begin{quote}
\it A \emph{ transient-all blame trail} is a sequence of scenarios
$\conf_0,...\conf_n$ of a program $\system$ where for all $0 \leq i \leq n - 1$,
$\conf_i \subset \conf_{i+1}$ and ~$\conf_{i+1} \setminus \conf_i = \mblame{\system,
\conf_i}$ where $\mblame{\system, \conf}$ denotes the set of components that $\conf$
blames under transient semantics.
\end{quote}
\noindent
and
\begin{quote}
\it A \emph{transient-firstborn blame trail} is a sequence of scenarios
$\conf_0,...\conf_n$ of a program $\system$ where for all $0 \leq i \leq n - 1$,
$\conf_i \subset \conf_{i+1}$ and $\conf_{i+1} \setminus \conf_i =
\first{\mblame{\system, \conf_i}}$ where $\first{\mblame{\system, \conf}}$ is the
first component transient adds to the blame set for $\conf$.
\end{quote}
The definitions of \emph{transient erasure trails} and \emph{transient random
trails} are analogous to those for natural erasure trails and natural random trails,
respectively.

We also use these four forms of blame trails to define when transient-all and
transient-firstborn add value to transient over exceptions and pure chance.
\begin{quote}
\it 
Transient all blame adds value to transient for a program $\system$ over
  pure chance and transient exceptions iff transient all blame 
  is more successful than both transient random and transient exception.

@@

\begin{figure}[ht]
\center
{\begin{tabular}{l|c|c|c}
%% ---------------------------------------------------------------------------------------------------------------
                        & {\bf Natural}        & {\bf Transient}          & {\bf Erasure} \\ \hline 
%% ---------------------------------------------------------------------------------------------------------------
{\bf Blame}             &       $Q_1$          &                          &               \\
{\bf All blame}         &                      &     $Q_2$                &               \\
{\bf Firstborn blame}   &                      &     $Q_3$                &               \\
{\bf Exception}         &       $Q_1$/$Q_*$    &     $Q_2$/$Q_3$/$Q_*$    &      $Q_*$    \\
{\bf Random}            &       $Q_1$/$Q_*$    &     $Q_2$/$Q_3$/$Q_*$    &      $Q_*$    \\
\end{tabular}}
  \caption{ Experimental questions and rational programmer modes.}
  \label{fig:experiment-outline}
\end{figure}

@@

In detail we answer $Q_1$ for a mutant by creating trails for each natural blame,
natural exception and natural random trail in the scenario lattice of the mutant. A
log reports on success and failure endings.  The first step is to construct the
mutant's scenario lattice and determine the scenario that can detect the fault in
the program. This can happen either because Typed Racket's type checker outright
rejects the scenario or because natural Typed Racket's runtime performs a type check
that fails. In either case, each of these scenarios is the root of a trail. Our test
bed extends these trails according to the natural-blame programmer.  If no scenarios
can be added to the trail---a trail may the top of the finite lattice---we check if
the last scenario of the trail type checks or not. If it does, this natural-blame
trail is successful, otherwise it is failing. We repeat the same process for the
roots two more times; once following the natural-exception programmer and once the
natural-random programmer. Finally, based on the collected data, we calculate the
success/failure results for the trails to determine whether natural blame is more
successful than natural exceptions and pure chance and we obtain the answer for
$Q_1$.  The process is analogues for the other questions employing the respective
modes of the rational programmer.
